---
title: "R Functions for Stardog"
author: "Brendan Newlon"
date: "12/1/2021"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
source_packages = function(pkgs, lib = .libPaths()[[1]], repos='http://cran.us.r-project.org') {
  new.pkgs = pkgs[!(pkgs %in% installed.packages()[, "Package"])]
  if (length(new.pkgs)){
    install.packages(new.pkgs, dependencies = T, lib = lib, repos=repos)
    }
  sapply(pkgs, require, character.only = T)
}

source_packages(c("knitr", "keyring", "SPARQL", "magrittr", "readr", "stringr", "rlang", "httr", "jsonlite"))
knitr::opts_chunk$set(echo = TRUE)
as.df = function(x)as.data.frame(x,strings.as.factors=F)
`%notin%` =Negate(`%in%`)
fix_json_encoding = function(x){x = gsub("\\\\u003d", "=" , x ) ; x }
```
# Start/Stop a Stardog server

These require that you have added the location of your Stardog /bin
directory to the system's PATH environment variable. (This version is
only for servers on localhost).

-   `start_stardog()`

-   `stop_stardog()`

```{r server, echo=FALSE, include=F}
start_stardog = function(server = "localhost"){
  if(server == "localhost"){  
    cat("Starting Stardog server on Localhost. \nThis may take a couple of minutes...\n") ; 
    stdout = suppressWarnings(
      system2(command = "cmd", args = c("/c", "stardog-admin server start"), wait = F, invisible = F, minimized = F, stdout = T, timeout = 10) 
    ) ; 
    cat("Started!\n")
    cat(stdout) 
  }## Remote handling TBD
}
stardog_start = function(){start_stardog}
# start_stardog()

##___________________## 
stop_stardog = function(server = "localhost"){
  if(server == "localhost"){ 
    cat("Stopping Stardog server on Localhost...\n") ; stdout = system2(command = "stardog-admin", args = c("server stop"), wait = F, invisible = F, minimized = F, stdout = T) ; 
    cat(stdout) 
  } ## Remote handling TBD
}
stardog_stop = function(){stop_stardog}
# stop_stardog()
```

```{r}
start_stardog()
```

# Secure credential handling

The function `handle_keys()` works behind the scenes to use the secure
credential store on your OS for connection details including passwords.

You'll be prompted to enter your password the first time you connect to
an endpoint with your username, but after that, your OS will manage
authentication silently. The function provides the option to delete
stored credentials from the system's secure credential store at any
time.

```{r echo=FALSE, include=F}
handle_keys = function(con_service = con_service, Username = "", clear_keys = F, as_password = F){
  # Establish username 
  if (!exists("Username") || Username == "" || is_empty(Username)) { 
    Username <- readline(prompt="Enter username: ") 
  } 
  if (!exists("Username") || Username == "" || is_empty(Username)) { 
    stop("Please try again with a valid username. ") 
  } 
  if(clear_keys){ 
    if (con_service %in% key_list(con_service)$service)  { 
      # Delete password 
      key_delete(service = con_service, username = Username) 
    }  
  } 
  # Ask for password if not already stored 
  if (!con_service %in% key_list(con_service)$service || !Username %in% key_list(con_service)$username)  { 
    if(Username == "anonymous"){
      key_set_with_value(service = con_service, username = Username,password = "anonymous") } else {
    # Store pasword securely 
    key_set(service = con_service, username = Username) 
    }
  }  
  if(as_password){return(key_get(service = con_service, username = Username))}
}
```

# Query Stardog

`stardog()` is a wrapper function that uses the SPARQL library to query
a Stardog endpoint.

```{r stardog, echo=FALSE}
# Note: This is a work-in-progress, please ignore the graph argument and specify FROM NAMED or GRAPH within the query text instead
stardog = function(
  server = "My_Localhost", 
  endpoint = "http://localhost:5820", 
  db = "stardog-tutorial-music", 
  graph = "", ## ignore this arg
  query = 'select * {?s ?p ?o} limit 10', 
  reasoning = F,
  Username = "admin", 
  include_namespaces = F, 
  clear_keys = F
){
  con_service = paste0("stardog_",server)
  handle_keys(con_service = con_service, Username = Username, clear_keys = clear_keys, as_password = F)
  sd_url = paste0(endpoint, paste0("/", db,graph, "/query/") %>% gsub("//","/",.))
  if(reasoning){sd_url = paste0(sd_url, "reasoning/")}
  sparql.result <- SPARQL(
    url = sd_url,
    query =  query,
    curl_args=c('userpwd'=paste0(Username,':',key_get(service = con_service, username = Username)))
  ) 
  if(!include_namespaces) { sparql.result = sparql.result %>% .[["results"]]}
  sparql.result
}
```

#### Example: Query Stardog on localhost

You'll be prompted for a password the first time, but successive queries
from the same username to the same endpoint will retrieve the password
from your system's secure credential store.

-   **The default password is `admin`**

```{r}
results = stardog(
  query = 'select * {?s ?p ?o} limit 5',
  server = "My_Localhost",
  endpoint = "http://localhost:5820",
  db = "stardog-tutorial-music",
  Username = "admin"
)
results
```



#### Example: Query Stardog Cloud in a SERVICE query while connected to localhost

1. Before starting your localhost Stardog server, create a file called services.sdpass in the folder specified by the STARDOG_HOME environment variable. Assuming the file/folder will only be readable by your user account, the file should contain the following text, substituting your username/password:         
  something.stardog.cloud:5820:*:YourUsername:YourPassword
2. Start the localhost Stardog server and run your query:

```{r}
results = stardog(
  query = '
    SELECT * {
      SERVICE <https://ucb-poc.stardog.cloud:5820/ucb/query> {
          ?s ?p ?o .
      }
    } 
    limit 10',
  server = "My_Localhost",
  endpoint = "http://localhost:5820",
  db = "stardog-tutorial-music",
  Username = "admin"
)
results
```


### Querying Stardog in shorthand

`stardog_()` -- note the underscore-- is a convenience wrapper function,
eg., for querying repeatedly within a function or to simplify the code
required to make repeated queries on the same environment. It calls the
function `stardog()` with the expectation that all the required
`stardog()` function arguments other than the query are already defined
and available in the caller environment.

```{r include=F, echo=F}
stardog_ = function(q = query, d = db, g = graph, U = Username, e = endpoint){
  x = stardog(query = q, db = d, graph = g, Username = U, endpoint = e ) ; x }
```

Define variables:

```{r}
endpoint = "http://localhost:5820"
db = "stardog-tutorial-music"
graph = ""
Username = "admin"
```

Query away!

```{r}
stardog_("select distinct ?song ?writer {?s a :Song ; :name ?song ; :writer ?w . ?w :name ?writer} order by desc(?writer) limit 5")
```

```{r}
stardog_("select ?album ?artist {?s a :Album; :name ?album; :artist ?a. ?a :name ?artist} order by ?s limit 3")
```

### Query Stardog from a Shiny app

The `stardog_shiny()` function adapts the function for a Shiny app that
has input fields for username and password, so it doesn't add or look
for credentials in the system's secure credential store.

```{r echo=F, include=F}
stardog_shiny = function(
  server = "My_Localhost", 
  endpoint = "http://localhost:5820", 
  db = "", 
  graph = "",
  query = 'select * {?s ?p ?o} limit 10', 
  reasoning = F,
  Username = input$username_input, ## Shiny app input - update for the id in your app
  Password = input$password_input,
  include_namespaces = F, 
  clear_keys = F
){
  sd_url = paste0(endpoint, paste0("/", db,graph, "/query/") %>% gsub("//","/",.))
  if(reasoning){ sd_url = paste0(sd_url, "reasoning/") }
  sparql.result = SPARQL(url = sd_url, query = query,curl_args = c('userpwd'= paste0(Username,':',Password)) ) 
  if(!include_namespaces) { sparql.result = sparql.result %>% .[["results"]] }
  sparql.result
}
```

# Send Stardog commands over HTTP

The `stardog_http()` function lets you send commands to Stardog via
HTTP.

```{r echo=F,include=F}
stardog_http = function( 
  query = "namespaces",  
  endpoint = "http://localhost:5820",
  db = "",
  graph = "", 
  Username = "admin",   
  clear_keys = F, 
  httr_method = "GET", # or "POST" etc.
  body = FALSE,
  assign_temp_results = F # used for debugging, to see exactly what was returned by the server
){ 
  con_service = "stardoghttp"
  handle_keys(con_service = con_service, Username = Username, clear_keys = clear_keys, as_password = F)
  sd_url = paste0(endpoint, paste0("/", db, "/", query) %>% gsub("//","/",.))
## Execute the query
  response = eval(parse(text= paste0("httr::",httr_method)))(
    url = sd_url %>% URLencode(), 
    add_headers(
      Authorization = paste0("Basic ", base64_enc(paste0(Username,":", key_get(service = con_service, username = Username) )) ),  
      accept = "text/html,application/xhtml+xml,application/xml;q=0.9,image/avif,image/webp,image/apng,*/*;q=0.8,application/signed-exchange;v=b3;q=0.9", 
      `accept-encoding` = "gzip, deflate, br" 
    ) , # end headers
    body = body
  ) 
  cat("status: ", response$status_code, "\n") 
  assign("Last.status", response$status_code %>% as.character, envir = .GlobalEnv)
  results =  response$content %>% rawToChar()  
  if(assign_temp_results){
    assign("temp_results", results, envir = .GlobalEnv) 
  }
  if(results != ""){results = results %>% fromJSON() 
  df = results[[1]] %>% as.df }
  df 
} 
```

Example: Get the namespaces and prefixes that are defined in the
**`covid19nyt`** database of our **`Stardog Cloud Express`** endpoint.

```{r}
stardog_http(
  query = "namespaces",
  endpoint = "https://express.stardog.cloud:5820",
  db = "covid19nyt",
  Username = "anonymous"
  )
```

### Shorthand HTTP calls

`stardog_http_()` --note the underscore at the end-- is a shorthand
wrapper function for convenience, like the `stardog_()` function above.

```{r echo=F, include=F}
stardog_http_ = function(q = query, d = db, g = graph, U = Username, e = endpoint, m = httr_method, b = body){
  x = stardog_http(query = q, db = d, graph = g, Username = U, endpoint = e, httr_method = m, body = b )
  x
}
```

### Stardog HTTP calls from a Shiny app

The `stardog_http_shiny()` function adapts the function for a Shiny app
that has input fields for username and password.

```{r echo=F, include=F}
stardog_http_shiny = function( 
  query = "namespaces",  
  endpoint = "",
  db = "",
  graph = "", 
  Username = input$username_input, ## example of retrieving Shiny user input
  Password = input$password_input,
  httr_method = "GET", # or "POST" etc.
  body = FALSE,
  assign_temp_results = F # For debugging, to see exactly what was returned by the server
){ 
  sd_url = paste0(endpoint, paste0("/", db, "/", query) %>% gsub("//","/",.))
  response = eval(parse(text= paste0("httr::",httr_method)))(
    url = sd_url %>% URLencode(), 
    add_headers( 
      Authorization = paste0("Basic ",  base64_enc(paste0(Username,":", Password ))   ),  
      accept = "text/html,application/xhtml+xml,application/xml;q=0.9,image/avif,image/webp,image/apng,*/*;q=0.8,application/signed-exchange;v=b3;q=0.9", 
      `accept-encoding` = "gzip, deflate, br" 
    ) , # end headers
    body = body
  ) 
  cat("status: ", response$status_code, "\n") 
  assign("Last.status", response$status_code %>% as.character, envir = .GlobalEnv)
  results =  response$content %>% rawToChar()  
  if(assign_temp_results){
    assign("temp_results", results, envir = .GlobalEnv) 
  }
  if(results != ""){results = results %>% fromJSON() 
  df = results[[1]] %>% as.df }
  df 
} 
```

# Create a Database

```{r echo=F, include=F}
stardog_create_db = function(
  db_name, 
  endpoint = "http://localhost:5820"
  ){
  r = stardog_http(
    endpoint = endpoint,
    httr_method = "POST", 
    query = 'admin/databases', 
    body = list(root = paste0('{"dbname": "',db_name,'"}'))
  ) 
  db_status = "" ; stardog_message = "";
  if(Last.status == "201"){db_status = paste("Database",db_name, "created!\n")}
  if(Last.status == "400"){db_status = paste("Database",db_name, "could not be created! Does it already exist?\n")
    stardog_message = paste("Stardog says:",r$x[1])
  }
  return(c(db_status, "\n", stardog_message))
}
```

Example (not run)

```{r eval=F}
stardog_create_db("cars", endpoint = "http://localhost:5820")
```

# A supporting function for cURL

`curlr()` is an alternative to `stardog_http()` and works behind the
scenes for some http operations.

```{r echo=F, include=F}
## Assume the following would succeed from the system console/terminal:
## curl -u admin:admin -X POST -F name=@data/namespaces.ttl http://localhost:5820/cars/namespaces  

curlr = function(
  ## ---- OPTION 1: directly enter curl_statement as string:
  curl_statement = "",
  ## ---- OPTION 2: construct the curl_statement from variables:
  url = "",
  endpoint = "http://localhost:5820",
  db = "",
  Username = "",
  myAuth = "", # eg. "admin:admin",  
  VERB = "POST",
  query = "",
  payload = "",
  error_log = file.path("logs", "curlr_last_error.txt"),
  ##---- more options
  return_json = FALSE, # default export as list
  show_progress = FALSE,
  show_response = FALSE, 
  assign_outcome = FALSE, 
  assign_to_env = caller_env()
){
  if(!file.exists(error_log)){
    log_dir = gsub("/[^/]*$", "",error_log)
    dir.create(file.path(log_dir), showWarnings = FALSE) 
    file.create(error_log) 
  }
  myUrl = if(url == "") {paste0(endpoint,if(db!="")paste0("/",db), "/", query)} else {url}
  if(myAuth!="")myAuth = paste0("-u ", gsub("-u ", "", myAuth))
  if(curl_statement == ""){
    if(myAuth == "" || Username != ""){
  con_service = "stardoghttp"
  # # Secure key handling using system key store
  handle_keys(con_service = con_service, Username = Username)
  myAuth = paste0(Username,":",key_get(service = con_service, username = Username))
  }
    curl_statement = paste0("curl ",myAuth," -X ",VERB," ",payload," ",myUrl)}
  ## Execute the cURL
  x = system2(command = "cmd", input = curl_statement, stdout = T, stderr = error_log)
  outcome = list(
    preamble = x[1:3], # Microsoft copyright statement etc
    command = x[4], 
    progress = if(x[5] == "  % Total    % Received % Xferd  Average Speed   Time    Time     Time  Current"){ x[5:7] ; had_progress = T} else {FALSE; had_progress = F}, 
    response = if(had_progress){ x[8:(length(x)-1)] } else {x[5:(length(x)-1)]}
  )
  if(had_progress && show_progress){
    cat(paste(outcome$progress , collapse = " \n"))
  }
  response = paste(outcome$response , collapse = " \n") %>% fix_json_encoding()
  if(response != ""){
  response = if(return_json){response}else{fromJSON(response)}
  if(show_response){
    if(return_json){
      cat(paste(outcome$response , collapse = " \n") %>% fix_json_encoding())
    } else {      print(response)    }
  }
  if(assign_outcome){assign("curlr_outcome", outcome, envir = assign_to_env)}
  return(response)
  } 
}
```

### Option 1: use `curlr()` to execute a cURL statement provided as a string

Example: get the namespaces registered in a database

```{r}
curlr(curl_statement = "curl -u anonymous:anonymous -X GET https://express.stardog.cloud:5820/covid19nyt/namespaces")
```

Example: using `curlr()` to get JSON results

```{r}
curlr(curl_statement = "curl -u anonymous:anonymous -X GET https://express.stardog.cloud:5820/covid19nyt/namespaces", return_json = T) 
```

### Option 2: construct and execute the cURL statement

Example: how to add namespaces defined in a file to a Stardog database
(not run)

```{r eval=F}
namespaces_file = file.path("data", "namespaces.ttl")

x = curlr(
  endpoint = "http://localhost:5820",
  db = "cars",
  Username = "admin",
  VERB = "POST",
  query = "namespaces",
  payload = paste0("-F name=@",namespaces_file)
)
```

# Add namespaces to a database

`stardog_add_namespaces()` is a dedicated function to do the same thing.
It also operates behind the scenes to automatically add namespaces when
you import CSV data. The input_file can be in either turtle or sms
format.

```{r echo=F, include=F}
stardog_add_namespaces = function(
  endpoint = "http://localhost:5820",
  db = "",
  Username = "admin",
  input_file,
  ...
)
{
  con_service = "stardoghttp"
  # # Secure key handling using system key store
  handle_keys(con_service = con_service, Username = Username)
  myAuth = paste0(Username,":",key_get(service = con_service, username = Username))
  ns_db = db
  namespaces = input_file %>%
    read_lines()
  sms_namespaces = namespaces %>%  # if it's an SMS file
    .[str_detect(tolower(.) ,"^[ ]*prefix " )] %>% paste("@",., " .",sep = "", collapse = " ") %>% gsub("^@ .$", "",.)
  rdf_namespaces = namespaces %>%  # if it's an RDF file
    .[str_detect(tolower(.), "^[ ]*@prefix ")] %>% paste(sep = " ", collapse = " ")
  namespaces2 = paste(sms_namespaces,rdf_namespaces, sep =" ") # either works.
  new_namespaces_file = file.path("data","namespaces.ttl")
  if(file.exists(new_namespaces_file)){file.remove(new_namespaces_file) %>% invisible()}
  write_file(namespaces2, new_namespaces_file)
  # send the cURL
  namespaces_file = new_namespaces_file
  r = curlr(
    db = ns_db,
    VERB = "POST",
    query = "namespaces",
    payload = paste0("-F name=@",namespaces_file),
    show_response = F,
    myAuth = myAuth,
    ...
  )
  file.remove(new_namespaces_file) %>% invisible()
  r
}
```

Example: adding namespaces to a database (not run)

```{r eval=F}
stardog_add_namespaces(db = "cars", input_file = file.path("data", "cars_mappings.sms"))
```

# Import data from a CSV file and mapping

```{r echo=F, include=F}
stardog_virtual_import = function( 
  endpoint = "http://localhost:5820", 
  db, 
  Username = "admin", 
  input_file_type = "DELIMITED", 
  input_file, mapping_file 
  ){ 
  con_service = "stardoghttp" 
  ## Secure key handling using system key store 
  handle_keys(con_service = con_service, Username = Username) 
  myAuth = paste0(Username,":",key_get(service = con_service, username = Username))
  ## create the db if it doesn't exist yet 
  db_list = stardog_http(query = "admin/databases",endpoint = endpoint,Username = Username) 
  db_result = "Database exists!"
  if(isTRUE(db %notin% db_list$x)) {db_result = stardog_create_db(db) }
  ## perform the virtual import
  my_curl = paste0('curl -u ',myAuth,' -F "database=',db,'" -F "mappings=<',mapping_file,'" -F "input_file_type=',input_file_type,'" -F "input_file=<',input_file,'" ',endpoint,'/admin/virtual_graphs/import' ) 
  curlr(curl_statement = my_curl ) 
  ## automatically add namespaces found in the imported data 
  r = stardog_add_namespaces(endpoint = endpoint,db=db, input_file = mapping_file, Username = Username) 
  return(c(db_result, r))
}
```

#### Example: Use virtual import to load data from a CSV file into a Stardog database

By default, a virtual import with this function will also automatically

-   create the database if it doesn't already exist, and

-   add any namespaces referenced in the mapping file to the database so
    you won't have to define them before using them in SPARQL queries

```{r}
stardog_virtual_import( 
  endpoint = "http://localhost:5820", 
  Username = "admin", 
  db = "cars", 
  input_file_type = "DELIMITED", 
  input_file = file.path("data","cars.csv"), 
  mapping_file = file.path("data", "cars_mappings.sms") 
  )
```

# Epilogue

Other functions are in development, so stay tuned!
